Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-28 paper/boneyard.tex

% Supplemental materials

>>>>

Despite the growing popularity of more sophisticated deep learning-based embedding models, here we leverage latent dirichlet allocation (i.e., topic modeling) to embed episode and recall text.  This decision was motivated by several factors.  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Second, words can mean different things in different contexts (e.g. ``bat'' may be the act of hitting a baseball, the object used for that action, or as a flying mammal).  Topic models are robust to this, allowing words to exist as part of multiple topics.  Last, topic models provide a straightforward means of recovering the weights for the particular words comprising a topic, enabling straightforward interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as the Universal Sentence Encoder, GPT-2, and GPT-3 offer context-sensitive encoding of text passages, but the encoding space is complex and non-linear, and thus recovering the original words used to fit the model is not straightforward. However, it is worth pointing out that our general framework is divorced from the particular choice of language model. Moreover, many of the aspects of our framework could be swapped out for other choices. For example, the language model, the timeseries segmentation model and the episode-recall matching function could all be customized to suit a particular question space or application. Indeed for some questions, recovery of the particular words used to describe a memory may not be necessary, and thus other text-modeling approaches (including the deep learning-based embedding models described above) may be preferable. Future work will explore the influence of particular model choices on the framework's efficacy.

In extending classical free recall analyses to our naturalistic memory framework, we recovered two patterns of recall dynamics central to list-learning studies: a heightened probability of initiating recall with the first presented ``item'' (in our case, episode events; Fig.~\ref{fig:list-learning}A) and a strong bias toward transitioning from recalling a given event to recalling the one immediately following it (Fig.~\ref{fig:list-learning}B).  However, equally noteworthy are the typical free recall results \textit{not} recovered in these analyses, as each highlights a fundamental difference between the list-learning paradigm and naturalistic memory paradigms like the one employed in the present study.  The most noticeable departure from hallmark free recall dynamics in these findings is the apparent lack of a serial position effect in Figure \ref{fig:list-learning}C, which instead shows greater and lesser recall probabilities for events distributed across the episode.  Stimuli in free recall experiments most often comprise lists of simple, common words, presented to participants in a random order. \citep[In fact, numerous word pools have been developed based on these criteria; e.g.,][]{FrieEtal82}.  These stimulus qualities enable two assumptions that are central to word list analyses, but frequently do not hold for real-world experiences.  First, researchers conducting list-learning studies may assume that the content at each presentation index is essentially equal, and does not possess attributes that would render it, on average, more or less memorable than others.  Such is rarely the case with real-world experiences or experiments meant to approximate them, and the effects of both intrinsic and observer-dependent factors on stimulus memorability are well established \citep[for review see][]{ChunTurk07, ByliEtal15, TyngEtal17}.  Second, the random ordering of list items ensures that (across participants, on average) there is no relationship between the thematic similarity of individual stimuli and their presentation positions---in other words, two successively presented items are no more likely to be highly semantically similar than they are to be highly dissimilar.  In most cases, the exact opposite is true of real-world episodes.  Our internal thoughts, our actions, and the physical state of the world around us all tend to follow a direct (often causal) progression.  As a result, each moment of our experience tends to be inherently more similar to surrounding moments than to those in the distant past or future.  Memory literature has termed this strong temporal autocorrelation ``context,'' and in various media that depict real-world events (e.g., movies or written stories), we recognize it as a \textit{narrative structure}.  While a random word list (by definition) has no such structure, the logical progression between ideas and actions in a naturalistic stimulus prompts the rememberer to recount presented events in order, starting with the beginning.  This tendency is reflected in our findings' second departure from typical free recall dynamics: a lack of increased probability of first recall for end-of-sequence events (Fig.~\ref{fig:list-learning}A).

Because they disregard presentation order-dependent variability in the stimulus content, analyses such as those in Figure \ref{fig:list-learning} enable a more sensitive analysis of presentation order-dependent temporal dynamics in free recall. Yet by the same token, they paint a wholly incomplete picture of memory for naturalistic episodes.  In an attempt to address this shortcoming, we have developed a framework in the present study that characterizes the explicit semantic content of the stimulus and subsequent recalls.  However, sensitivity to stimulus and recall content introduces a new challenge: distinguishing between levels of recall quality for a stimulus (e.g., an event) that is considered to have been ``remembered.''  When modeling memory in an experimental setting, recall quality for individual events is often cast as binary (e.g., a given list item was simply either remembered or not remembered).  Various models of memory (e.g., \citealp{Yone02}) attempt to improve upon this by including confidence ratings, rendering this binary judgement instead categorical.  To better evaluate naturalistic memory quality, we introduce a continuous metric (\textit{precision}), which reflects the level of completeness of a participant's recall for a feature-rich experience.  Additionally, recall quality for a single event is typically assessed independently from that for all other events (e.g., it is difficult to ``compare'' a participant's binary recall success for list item 1 to that of list item 10).  The second novel metric we introduce (\textit{distinctiveness}) is based on analyzing of the correlational structure of an individual's full set of recall events, and reflects the specificity of their memory for a single experienced event.  We find that both of these metrics relate to the overall number of episode events participants successfully recalled, and that our precision metric additionally relates to \cite{ChenEtal17}'s hand-annotated memory memory scores.

We did not find evidence that participants' average recall distinctiveness was related to their hand-annotated memory scores computed by~\cite{ChenEtal17}.  One possible explanation is that, in hand-scoring each participant's verbal recall for each of 50 (manually-deliniated) scenes, ``[a] scene was counted as recalled if the participant described any part of the scene''~\citep{ChenEtal17}.  In other words, both an extensive description of a scene's content and a brief mention of some subset of its content were (binarily) considered equally successful recalls.  By contrast, we identify the event structure in participants' recalls in an unsupervised manner, independent of the episode event-timeseries, prior to mapping between episode and recall content.  Our HMM-based event-segmentation produces boundaries between timepoints where the topic proportions shift in a substantial way, and because a small handful of words is unlikely to contribute significantly to the topic proportions for any sliding window, such brief scene descriptions will most often not result in a sufficiently large shift in the resulting topic proportions for the HMM to identify an event boundary.  Instead, they will be grouped with a neighboring event, consequently lowering that event's distinctiveness score and by extension, the participant's overall distinctiveness score.  This is in essence the qualitative difference between distinctive and indistinctive recall, and reflects the comparison shown in Figure \ref{fig:distinctiveness-detail}C.  Intriguingly, prior studies show that pattern separation, or the ability to cleanly discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore whether and how these metrics compare between cognitively impoverished groups and healthy controls.

In the analyses outlined in Figure \ref{fig:brainz}, we identified two networks of brain regions whose responses during episode viewing were consistent with the temporal structure of the episode and recall topic trajectories, respectively. The network identified by the episode trajectory analysis included the ventromedial prefrontal cortex, left anterior temporal lobe, superior parietal and dorsal anterior cingulate cortex. The network from the episode-recall trajectory analysis also included the ventromedial prefrontal and superior parietal cortices, in addition to the posterior medial cortex (PMC) and the inferior temporal regions. Notably, \cite{ChenEtal17} also observed the PMC in a number of analyses including one that searched for regions whose activity patterns during encoding were reinstated during free recall. The PMC has been consistently identified in studies involving stimuli with meaningfully structured events~\citep{CohnRang17}. Further, the PMC is part of the "posterior medial" system, a network of brain regions thought to represent situation models~\citep{ZackEtal07} in support of memory, spatial navigation and social cognition~\citep{RangRitc12}. Given that we constructed our episode-recall searchlight model to capture temporal structure in the episode's semantic content (and how one's later recall aligns with that structure), we speculate that the PMC may play a role in constructing mnemonic events from meaningfully structured experiences.

Decoding the associated significance maps with \texttt{Neurosynth} revealed two intriguing results.  First, the top 10 terms returned for the episode-driven searchlight significance map were centered around themes of language and semantic meaning (Fig.~\ref{fig:brainz}).  In other words, the voxels identified as more reflective of the episode content's temporal structure (i.e., voxels with lower permutation correction-derived $p$-values), as defined by our model, were most likely to be reported as active in studies focused on the the neural underpinnings of semantic processing.  This finding is interesting, as our model specifically captures the temporal structure of the episode's \textit{semantic} content (e.g., as opposed to that of the visual, auditory, or affective content).  This suggests that the network of structures displayed in Figure \ref{fig:brainz}C may play a roll in processing the evolving semantic content of ongoing experiences.

Our second searchlight analysis identified a partially overlapping network of regions (Fig.~\ref{fig:brainz}D) whose patterns of activity as participants viewed the episode reflected the idiosyncratic structure of each individual's later recalls.  The associated significance map yielded a set of \texttt{Neurosynth} terms that primarily reflected names of specific structural regions (such as ``thalamus,'' ``anterior insula,'' ``anterior cingulate'' and ``inferior frontal''; Fig.~\ref{fig:brainz}).  Interestingly, these regions share membership in a common, large-scale functional network (termed the ``salience network'') involved in detecting and processing affective cues.  In particular, the latter three regions have been implicated in functions relevant to assigning personal meaning to an experience, including: ascribing subjective value to raw, sensory input \citep{MedfCrit10}; modulating semantic and phonological processing in response to personally salient stimuli \citep{KellEtal07b}; and directing and reallocating attention and working memory resources towards the most relevant stimuli \citep{MenoUddi10}.  This suggests that the network of structures displayed in Figure \ref{fig:brainz}D may be play a roll in transforming and restructuring ongoing experiences through the lens of one's prior experience and subjective emotions as they are encoded in memory.

%

Supplemental materials

### 2020-08-28 paper/main.tex

Explicitly modeling the dynamic content of a naturalistic stimulus and participants' memories enabled us to connect the present study of naturalistic recall with an extensive prior literature on using list-learning paradigms to study memory~\citep[for review see][]{Kaha12}, as in Figure~\ref{fig:list-learning}.  We found some similarities between how participants in the present study recounted a television episode and how participants typically recall memoized random word lists.  However, our broader claim is that word lists miss out on fundamental aspects of naturalistic memory more like the sort of memory we rely on in everyday life.  For example, there are no random word list analogs of character interactions, conceptual dependencies between temporally distant episode events, the sense of solving a mystery that pervades the \textit{Sherlock} episode, or the myriad other features of the episode that convey deep meaning.  Nevertheless, each of these properties affects how people process the ongoing episode as they are watching it, and how they remember it later.  The overarching goal of the present study is to characterize how the rich dynamics of the episode affect the rich behavioral and neural dynamics of how people remember it.

>>>>

Explicitly modeling the dynamic content of a naturalistic stimulus and participants' memories enabled us to connect the present study of naturalistic recall with an extensive prior literature that has used list-learning paradigms to study memory~\citep[for review see][]{Kaha12}, as in Figure~\ref{fig:list-learning}.  We found some similarities between how participants in the present study recounted a television episode and how participants typically recall memorized random word lists.  However, our broader claim is that word lists miss out on fundamental aspects of naturalistic memory more like the sort of memory we rely on in everyday life.  For example, there are no random word list analogs of character interactions, conceptual dependencies between temporally distant episode events, the sense of solving a mystery that pervades the \textit{Sherlock} episode, or the myriad other features of the episode that convey deep meaning.  Nevertheless, each of these properties affects how people process the ongoing episode as they are watching it, and how they remember it later.  The overarching goal of the present study is to characterize how the rich dynamics of the episode affect the rich behavioral and neural dynamics of how people remember it.
